---
title: "UD651 - Problem Set 4"
author: "Ivan Heneghan"
date: "16 January 2016"
output: html_document
---

This is the project report for UD651 - Problem Set 4.

Load relevant libraries:

```{r}
library(data.table)
library(plyr)
library(dplyr, warn.conflicts = FALSE)
library(ggplot2)
library(gridExtra)
library(lubridate)
library(reshape2)
library(tidyr)
library(xlsx)
```

1. Scatterplot - price vs x

```{r}
ggplot(data = diamonds, aes(x = price, y = x)) + 
  geom_point() +
  xlab("Price") +
  ylab("X") +
  ggtitle("Price vs X")
```

2. Correlations - price vs. x, y, z

```{r}
with(diamonds, cor.test(price, x, method = "pearson"))
with(diamonds, cor.test(price, y, method = "pearson"))
with(diamonds, cor.test(price, z, method = "pearson"))
```

3. Scatterplot - price vs depth

```{r}
ggplot(data = diamonds, aes(x = price, y = depth)) + 
  geom_point() +
  xlab("Price") +
  ylab("Depth") +
  ggtitle("Price vs Depth")
```

4. Scatterplot - price vs depth adjustments

4a. Get range of price

```{r}
range(diamonds$depth)
```

4b. Plot

```{r}
ggplot(data = diamonds, aes(x = depth, y = price)) + 
  geom_point(alpha = 1/100) +
  scale_x_continuous(breaks = seq(43, 79, 2)) +
  xlab("Depth") +
  ylab("Price") +
  ggtitle("Price vs Depth")
```

4c. Correlation of price & depth

```{r}
with(diamonds, cor.test(price, depth, method = "pearson"))
```

5. Scatterplot - price vs carat

5a. Get range of price

```{r}
range(diamonds$carat)
```

5b. Plot

```{r}
ggplot(data = diamonds, aes(x = carat, y = price)) + 
  geom_point(stat = "summary", fun.y = quantile, fun.args = list(probs = .99)) +
  scale_x_continuous(breaks = seq(0.2, 5.01, 0.5)) +
  xlab("Carat") +
  ylab("Price") +
  ggtitle("Price vs Carat")
```

6. Scatterplot - price vs volume

6a. Create volume variable

```{r}
diamonds$volume <- diamonds$x * diamonds$y * diamonds$z
range(diamonds$volume)
```

6b. Scatterplot

```{r}
ggplot(data = diamonds, aes(x = volume, y = price)) + 
  geom_point() +
  scale_x_continuous(breaks = seq(0, 3841, 50)) +
  xlab("Volume") +
  ylab("Price") +
  ggtitle("Price vs Volume")
```

7. Scatterplot - removing outliers through subset

7a. Create subsetted data set

```{r}
diamondsReduced <- subset(diamonds, volume > 0 & volume <= 800)
```

7b. Correlation of this

```{r}
with(diamondsReduced, cor.test(price, volume, method = "pearson"))
```

7c. Adjustments

```{r}
ggplot(data = diamondsReduced, aes(x = volume, y = price)) + 
  geom_point(alpha = 1/10) +
  geom_smooth(method = "lm", color = "red") + 
  scale_x_continuous(breaks = seq(0, 3841, 50)) +
  xlab("Volume") +
  ylab("Price") +
  ggtitle("Price vs Volume")
```

8. Mean price by clarity

8a. Creation of the new data frame

```{r}
diamondsByClarity <- diamonds %>% 
  group_by(clarity) %>% 
  summarise(mean_price = mean(price),
            median_price = median(price),
            min_price = min(price),
            max_price = max(price),
            n = n()) %>% 
  arrange(clarity)
```

8b. Bar charts of mean price

```{r}
diamonds_by_clarity <- group_by(diamonds, clarity)
diamonds_mp_by_clarity <- summarise(diamonds_by_clarity, mean_price = mean(price))

diamonds_by_color <- group_by(diamonds, color)
diamonds_mp_by_color <- summarise(diamonds_by_color, mean_price = mean(price))

g1 <- ggplot(data = diamonds_mp_by_clarity, aes(x = clarity, y = mean_price)) + 
  geom_bar(stat="identity")
g2 <- ggplot(data = diamonds_mp_by_color, aes(x = color, y = mean_price)) + 
  geom_bar(stat="identity")
grid.arrange(g1, g2, ncol = 1)
```

8c. Looking at mean price/cut relationship

```{r}
diamonds_by_cut <- group_by(diamonds, cut)
diamonds_mp_by_cut <- summarise(diamonds_by_cut, mean_price = mean(price))
g3 <- ggplot(data = diamonds_mp_by_cut, aes(x = cut, y = mean_price)) + 
  geom_bar(stat="identity")
grid.arrange(g1, g2, g3, ncol = 1)
```

9. Gapminder revisited - look at link between mobile phone ownership & income per person/GDP per capita. 

9a. Go to right directory

9b. Import data

```{r}
cell <- read.csv("cell phone per 100.csv", 1, stringsAsFactors = FALSE)
gdp <- read.csv("GDPpercapitaconstant2000US.csv", 1, stringsAsFactors = FALSE)
```

9c. Check data

```{r}
str(cell)
head(cell)
str(gdp)
head(gdp)
```

9d. Clean up cell data

```{r}
names(cell)[1] <- c("country")
cell <- cell %>% gather(year, subs, -country)
cell$year <- gsub("X", "", cell$year)
```

9e. Clean up gdp data

```{r}
names(gdp)[1] <- c("country")
gdp <- gdp %>% gather(year, gdp, -country)
gdp$year <- gsub("X", "", gdp$year)
```

9f. Subset to a matching timeframe - 1981 to 2011

```{r}
cell <- subset(cell, year >= 1981 & year <= 2011)
gdp <- subset(gdp, year >= 1981 & year <= 2011)
```

9g. Merge data sets & clean NA's

```{r}
merged <- merge(cell, gdp, all = TRUE)
merged <- filter(merged, !is.na(merged$subs))
merged <- filter(merged, !is.na(merged$gdp))
```

9h. Check ranges

```{r}
summary(merged$subs)
summary(merged$gdp)
```

9i. Scatterplot using alpha & jitter shows a positive correlation - as a countrie's GDP increases, cell phone ownership increases.

```{r}
ggplot(aes(x = gdp, y = subs), data = merged) +
  geom_point(alpha = 1/10, position = position_jitter(h = 0)) +
  geom_smooth(method = "lm", color = "red") +
  scale_x_continuous() +
  scale_y_continuous() +
  labs(title = "GDP vs Cell Phones",
       x = "GDP per Capita",
       y = "Cell Phones per 100 People")
```

9j. Look at correlation to reinforce viewpoint

```{r}
cor.test(merged$gdp, merged$subs, method = "pearson")
```

9k. Look @ 95th percentiles & correlations to strip out massive outliers. Viewpoint still holds true

9ki. Look @ 95th percentile of GDP - plot & correlation

```{r}
ggplot(aes(x = gdp, y = subs), data = merged) +
  geom_point(alpha = 1/10, position = position_jitter(h = 0)) +
  xlim(0, quantile(merged$gdp, 0.95)) + 
  geom_smooth(method = "lm", color = "red")
```

```{r}
merged_gdp95th <- subset(merged, gdp <= quantile(gdp, 0.95))
cor.test(merged_gdp95th$gdp, merged_gdp95th$subs, method = "pearson")
```

9kii. Look @ 95th percentile of subs - plot & correlation

```{r}
ggplot(aes(x = gdp, y = subs), data = merged) +
  geom_point(alpha = 1/10, position = position_jitter(h = 0)) +
  ylim(0, quantile(merged$subs, 0.95)) +
  geom_smooth(method = "lm", color = "red")
```

```{r}
merged_subs95th <- subset(merged, subs <= quantile(subs, 0.95))
cor.test(merged_subs95th$gdp, merged_subs95th$subs, method = "pearson")
```

9kiii. Look @ 95th percentile for both - plot & correlation

```{r}
ggplot(aes(x = gdp, y = subs), data = merged) +
  geom_point(alpha = 1/10, position = position_jitter(h = 0)) +
  xlim(0, quantile(merged$gdp, 0.95)) + 
  ylim(0, quantile(merged$subs, 0.95)) +
  geom_smooth(method = "lm", color = "red")
```

```{r}
merged_95th <- subset(merged, gdp <= quantile(gdp, 0.95) & subs <= quantile(subs, 0.95))
cor.test(merged_95th$gdp, merged_95th$subs, method = "pearson")
```

9l. Add mean, median, percentile (25th, 75th) - ask Srdjan, as this doesn't work well. Too many data points in y?

```{r}
ggplot(aes(x = gdp, y = subs), data = merged) +
  geom_point(alpha = 1/10, position = position_jitter(h = 0)) +
  geom_line(stat = 'summary', fun.y = quantile, fun.args = list(prob = .25), color = 'purple') +
  scale_x_continuous() +
  scale_y_continuous() +
  labs(title = "GDP vs Cell Phones",
       x = "GDP per Capita",
       y = "Cell Phones per 100 People")
```

9m. Focus on a single year - 2011. 

```{r}
merged2011 <- subset(merged, year == 2011)
```

9n. Scatterplot & correlation for 2011 continues to reinforce positive relationship

```{r}
ggplot(aes(x = gdp, y = subs), data = merged2011) +
  geom_point(alpha = 1/10, position = position_jitter(h = 0)) +
  geom_smooth(method = "lm", color = "red") +
  scale_x_continuous() +
  scale_y_continuous() +
  labs(title = "GDP vs Cell Phones - 2011",
       x = "GDP per Capita",
       y = "Cell Phones per 100 People")
```

```{r}
cor.test(merged2011$gdp, merged2011$subs, method = "pearson")
```

9o. Look at top 10 countries & bottom 10 countries for 2011

```{r}
sorted2011 <- arrange(merged2011, desc(gdp))
top102011 <- head(sorted2011, n = 10)
bottom102011 <- tail(sorted2011, n = 10)
```

9p. Scatterplot & correlation for both shows positive correlation. However, much stronger correlation for bottom 10 - therefore, cell phone ownership as a factor of GDP much stronger in lower GDP countries (i.e. more cell phones owned by countires with lower GDP, and as GDP increases, cell phone ownership increases, at a faster rate in lower GDP countries.)

```{r}
ggplot(aes(x = gdp, y = subs), data = top102011) +
  geom_point(alpha = 1/10, position = position_jitter(h = 0)) +
  geom_smooth(method = "lm", color = "red") +
  scale_x_continuous() +
  scale_y_continuous() +
  labs(title = "GDP vs Cell Phones - 2011, Top 10 Countries by GDP",
       x = "GDP per Capita",
       y = "Cell Phones per 100 People")
```

```{r}
cor.test(top102011$gdp, top102011$subs, method = "pearson")
```

```{r}
ggplot(aes(x = gdp, y = subs), data = bottom102011) +
  geom_point(alpha = 1/10, position = position_jitter(h = 0)) +
  geom_smooth(method = "lm", color = "red") +
  scale_x_continuous() +
  scale_y_continuous() +
  labs(title = "GDP vs Cell Phones - 2011, Top 10 Countries by GDP",
       x = "GDP per Capita",
       y = "Cell Phones per 100 People")
```

```{r}
cor.test(bottom102011$gdp, bottom102011$subs, method = "pearson")
```